Variational Monte Carlo on Bosonic Systems

### 2018-04-06 project_1/article/article.tex
Added Haakon as reference.
Add section on initial spread

        \begin{enumerate}
            \item The system is initialised by a certain number $N$ of randomly
                generated positions, $\vf{r} = (\vf{r}_1, \dots, \vf{r}_N)$,
                which represent the particles. This allows us to evaluate the
                wavefunction at these points and compute the local energy
                $E_L(\vf{r})$.

            \item The initial configuration is changed by setting a new position
                for one of these particles. The particle is picked at random.

            \item A ratio, $q$, between the probability density of the new
                wavefunction and the previous probability density is computed.
                We accept the step if $q \geq 1$, else we accept the step with a
                probability $p$. The latter is done by drawing a random uniform
                number, $x \in [0, 1]$, and checking if $q > x$.  This
                acceptance probability decides if the particle move is rejected
                or accepted.  The particle is only allowed to move a
                predetermined step length.

            \item If the particle movement is accepted then the local energy is
                computed for the new system.

            \item Repeat steps until convergence and an optimum is reached.

@@

        The algorithm described above can be applied in an "exhaustive" search
        of the parameter space in order to find the optimal parameters.  Whether
        a proposed move is accepted or not is determined by a transition
        probability and the acceptance probability.  The strength of the
        algorithm lies in that the transition algorithm need not be known. For
        example, the simplest case is to accept the new state, i.e., the new
        position for the random walker, if the ratio
        \begin{align}
            q(\vf{r}_{i + 1}, \vf{r}_i)
            &=
            \frac{\left|\Psi_T(\vf{r}_{i + 1})\right|^2}
            {\left|\Psi_T(\vf{r}_{i})\right|^2},
        \end{align}
        where $\vf{r}_{i + 1}$ are all the positions at step $i + 1$, is greater
        than a uniform probability $p \in [0, 1]$.

        \subsubsection{Importance Sampling}
            A problem with the naïve Metropolis-Hastings sampling approach is
            that the sampling of position space is done with no regard for where
            we are likely to find a particle. This problem can be remedied
            through \emph{importance sampling}.  It is reasonable to assume that
            the particles we erratically scatter in space are prone to movement
            towards the peaks of the probability density as dictated by the wave
            function. Consider therefore the Fokker-Planck equation,
            \begin{align}
                \dpd[]{\Psi_T}{t}
                &=
                D \nabla\cdot
                \left(
                    \nabla
                    - \vf{F}
                \right) \Psi_T,
                \label{eq:fokker_planck}
            \end{align}
            which describes the evolution in time of a probability density
            function. In our case this is the trial wavefunction $\Psi_T$.
            Originaly an equation that models diffusion, we have a diffusion
            term $D$ and the drift force \autoref{eq:drift_force}. In our case
            the diffusion term $D$ is simply $1/2$ from the kinetic energy (in
            natural units).

>>>>

        \begin{enumerate}
            \item The system is initialised by a certain number $N$ of randomly
                generated positions, $\vf{r} = (\vf{r}_1, \dots, \vf{r}_N)$,
                which represent the particles\footnote{Also known as
                \emph{random walkers}.}. This allows us to evaluate the
                wavefunction at these points and compute the local energy
                $E_L(\vf{r})$.

            \item The initial configuration is changed by setting a new position
                for one of these particles. The particle is picked at random and
                is only allowed to move a predetermined step length.

            \item A ratio, $q$, between the probability density of the new
                wavefunction and the previous probability density is computed.
                We accept the step if $q \geq 1$, else we accept the step with a
                probability $p$. The latter is done by drawing a random uniform
                number, $x \in [0, 1]$, and checking if $q > x$.  This
                acceptance probability decides if the particle move is rejected
                or accepted.

            \item If the particle movement is accepted we compute the local
                energy for the new system. If not, we reset the system to the
                previous state and reuse the last local energy.

            \item Accumulate the local energy. Repeat the steps until
                convergence is reached or a set number of iterations has been
                done. When finished divide the accumulated energy by the number
                of cycles to compute the expectation value of the energy.

@@

        The algorithm described above can be applied in an "exhaustive" search
        of the parameter space in order to find the optimal parameters.  Whether
        a proposed move is accepted or not is determined by a \emph{transition
        probability}. The strength of the algorithm lies in that the transition
        algorithm need not be known. For example, the simplest case is to accept
        the new state, i.e., the new position for the random walker, if the
        ratio
        \begin{align}
            q(\vf{r}_{i + 1}, \vf{r}_i)
            &=
            \frac{\left|\Psi_T(\vf{r}_{i + 1})\right|^2}
            {\left|\Psi_T(\vf{r}_{i})\right|^2},
        \end{align}
        where $\vf{r}_{i + 1}$ are all the positions at step $i + 1$, is greater
        than a uniform probability $p \in [0, 1]$. This transition criteria
        results in what we call the brute force Metropolis-Hastings algorithm.
        By counting the number of accepted steps and dividing by the total
        number of Monte Carlo cycles we get the \emph{acceptance probability}.
        The acceptance probability helps in choosing a decent length for the
        proposed new step. If the acceptance probability is very low we can
        lower the step length to get more accepted steps and vice versa.

        \subsubsection{Initial distribution of particles}
            Prior to performing the Metropolis-Hastings algorithm we have to an
            initial distribution of the particles. As the algorithm is supposed
            to move particles in the right direction this distribution can be
            \emph{somewhat} arbitrary. There are however a few caveats to this
            approach.

            \begin{enumerate}
                \item The distribution is unphysical. This results in the
                    particles taking up positions which can yield bogus values
                    for the local energy. A remedy is to allow for some
                    \emph{thermalization steps}, i.e., scatter the particles and
                    run the algorithm for a set number of steps prior to
                    starting the actual sampling.
                \item The initial \emph{spread} of the particles is too small or
                    too large. The former can lead to an acceptance ratio of
                    zero if no steps are accepted due to overlap. This is
                    something that can happen to the interacting system as an
                    overlap between the particles makes the wavefunction
                    evaluate to zero thus rejecting all steps in the ratio test.
                    The latter can lead to the system never reaching a true
                    state as they are scattered too far away from the
                    interesting regions in the configuration space.
                \item The system imposes some restrictions on the distribution.
                    This can mean that we have to distribute the particles in a
                    certain way to make sure that we don't break the
                    restrictions. This resembles the former case when the spread
                    is too small for our systems.
            \end{enumerate}

            In the case of the interacting system we are looking at we have to
            take some measures to make sure that the initial spread does not
            break the Monte Carlo cycles. We therefore distribute all the
            particles on a grid with a distance larger than the diameter of the
            hard sphere before letting the system thermalize. After the
            thermaliztion we start the sampling.

        \subsubsection{Importance Sampling}
            A problem with the naïve Metropolis-Hastings sampling approach
            described above is that the sampling of position space is done with
            no regard for where we are likely to find a particle. This problem
            can be remedied through \emph{importance sampling}.  It is
            reasonable to assume that the particles we erratically scatter in
            space are prone to movement towards the peaks of the probability
            density as dictated by the wave function. Consider therefore the
            Fokker-Planck equation,
            \begin{align}
                \dpd[]{\Psi_T}{t}
                &=
                D \nabla\cdot
                \left(
                    \nabla
                    - \vf{F}
                \right) \Psi_T,
                \label{eq:fokker_planck}
            \end{align}
            which describes the evolution in time of a probability density
            function. In our case this is the trial wavefunction $\Psi_T$.
            Originaly an equation that models diffusion, we have a diffusion
            term $D$ and the drift force \autoref{eq:drift_force}. In our case
            the diffusion term $D$ is simply $1/2$ from the kinetic energy (in
            natural units).
